---
title: "Walkthrough: mixed data analysis"
output: 
  bookdown::html_document2:
    css: "style.css"
    number_sections: yes
    toc: true
    toc_depth: 2
    toc_float:
      collapsed: false
vignette: >
  %\VignetteIndexEntry{main-features}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r klippy, echo = FALSE}
# Create a button for copy pasting code chunk to clipboard
klippy::klippy(position = c("top", "right"))

```

```{r global_options, include = FALSE}
# Load knitr package to change options
library(knitr)

# Set default options for the chunks
opts_chunk$set(
    echo = TRUE,
    warning = FALSE,
    message = FALSE,
    comment = ""
)

# Set the width of the console output to a large number (for sidewise scroll)
options(width = 1e3)

```

In this vignette we explore how to use `GSPCR` for the analysis of a data set with a mix of measurement levels.
In certain real applications, the dependent variable might be recorded on a continuous numeric scale, while some or all independent variables might be recorded on a binary or even a nominal scale.
`GSPCR` supports any combination of these measurement levels.

```{r setup}
# Load R packages
library(gspcr)      # this package!
library(superpc)    # alternative comparison package
library(patchwork)  # combining ggplots

# Set a seed
set.seed(20230415)

```



```{r}
# Look at the helpfile
?GSPCRexdata

# Comment goal of code
X <- GSPCRexdata$X$mix
y <- GSPCRexdata$y$cont

```

# Continuous dependent variable, mix of predictors

To explore this scenario, we will run `gspcr` using the continuous dependent variable and the mix of continuous and categorical predictors in the example data set `GSPCRexdata`.
As a threshold measure, we cannot use the `normalized` bivariate association used in the original `superpc` R package which requires continuous dependent and independent variables.

```{r }
# Train the GSPCR model
out <- cv_gspcr(
  dv = GSPCRexdata$y$cont,
  ivs = GSPCRexdata$X$mix,
  fam = "gaussian",
  nthrs = 20,
  npcs_range = 1:5,
  K = 10,
  fit_measure = "F",
  thrs = "PR2",
  min_features = 1,
  max_features = ncol(X),
  oneSE = TRUE
)

# Use the plotting function
plot_output <- plot(
  x = out,
  y = "F",
  labels = TRUE,
  errorBars = TRUE,
  discretize = FALSE
)

# And plot
plot_output

```

# TL;DR, just give me the code!

```{r TLDR, ref.label = knitr::all_labels()[!knitr::all_labels() %in% c("global_options", "klippy")], echo = TRUE, eval = FALSE}
```

<!-- Remove white space generated by toc_float -->
<div class="tocify-extend-page" data-unique="tocify-extend-page" style="height: 0;"></div>